---
title: "Oʻahu Rainfall and Watershed Risk Assessment Over Time"
author: "Jaydon Laboy, Sano Tanna, Kyle-Jacob Gamiao"
format: html
editor: visual
---

## Library Statements:

```{r}
library("sf")
library("terra")
library("tidyverse")
library("lubridate")
library("exactextractr")
library("shiny")
library("rgbif")
library("RColorBrewer")
library("broom")
library("ggplot2")
library("patchwork")
library("leaflet")
library("units")
library("randomForest")
library("elevatr")
```

## Rainfall Data From 1990-2025

Loading and cleaning rainfall data.

```{r}
# List all rainfall GeoTIFFs (Change this path to where you have the data stored)
rain_base_path <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data/oahu_rain"

all_tifs <- list.files(rain_base_path, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter years
rain_files <- all_tifs[grepl("_(199\\d|200\\d|201\\d|202[0-5])_\\d{2}\\.tif$", all_tifs)]

# Checking if files loaded in properly
if(length(rain_files) == 0){
  stop("No TIF files found for 1990-2025. Check your folder or filenames!")
}

# Stack all the files
rain_stack <- rast(rain_files)

# Name each of the layers by YYYY_MM from filename
names(rain_stack) <- gsub(".*_(\\d{4}_\\d{2})\\.tif$", "\\1", basename(rain_files))
```

Monthly and annual computations

```{r}
# Monthly metrics
rain_metrics <- map_dfr(1:nlyr(rain_stack), function(i){
  r <- rain_stack[[i]]
  vals <- values(r)
  vals <- vals[!is.na(vals)]
  
  data.frame(
    month = names(r),
    mean_rain_mm = mean(vals),
    min_rain_mm  = min(vals),
    max_rain_mm  = max(vals),
    sd_rain_mm   = sd(vals),
    p10_rain_mm  = quantile(vals, 0.10),
    p90_rain_mm  = quantile(vals, 0.90)
  )
})
rain_metrics

# Annual means
annual_mean <- mean(rain_stack)
annual_mean
```

Annual Rainfall Map

```{r}
# Convert raster to data frame for ggplot
annual_df <- as.data.frame(annual_mean, xy=TRUE)
colnames(annual_df) <- c("x", "y", "rain_mm")

# Plot annual rainfall map
ggplot(annual_df, aes(x = x, y = y, fill = rain_mm)) +
  geom_raster() +
  scale_fill_gradientn(
    colors = c("#2E8B57", "#ADD8E6"),  # green → light blue
    name = "Rainfall (mm)",
    guide = guide_colorbar(
      barwidth = 1.5, barheight = 15,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  coord_equal() +
  labs(
    title = "Oʻahu Average Rainfall 1990–2025",
    subtitle = "Values in millimeters (mm) per month",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

Annual Mean Rainfall w/ Slider

```{r}
# Load Rainfall Data
rain_base <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data/oahu_rain"

all_tifs <- list.files(rain_base, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter monthly files like *_1990_03.tif
rain_files <- all_tifs[grepl("_(199\\d|200\\d|201\\d|202[0-5])_\\d{2}\\.tif$", all_tifs)]

if(length(rain_files)==0) stop("No rainfall rasters found!")

rain_stack <- rast(rain_files)

# Extract layer names YYYY_MM
names(rain_stack) <- gsub(".*_(\\d{4}_\\d{2})\\.tif$", "\\1", basename(rain_files))

# Extract Years and Compute Rasters
years <- as.integer(substr(names(rain_stack), 1, 4))
unique_years <- sort(unique(years))

annual_rain_rasters <- list()

for(y in unique_years){
  idx <- which(years == y)
  annual_rain_rasters[[as.character(y)]] <- mean(rain_stack[[idx]])
}

# Color Palette
all_vals <- unlist(lapply(annual_rain_rasters, values))

pal_rain <- colorNumeric(
  palette = c("#2E8B57", "#ADD8E6"), 
  domain  = all_vals,
  na.color = "transparent"
)

# Shiny UI
ui <- fluidPage(
  titlePanel("Oʻahu Annual Mean Rainfall (1990–2025)"),
  sidebarLayout(
    sidebarPanel(
      sliderInput(
        "year",
        "Select Year:",
        min = min(unique_years),
        max = max(unique_years),
        value = min(unique_years),
        step = 1,
        sep = ""
      )
    ),
    mainPanel(
      leafletOutput("rain_map", height = "600px")
    )
  )
)

# Shiny Server
server <- function(input, output, session){

  output$rain_map <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%
      addRasterImage(
        annual_rain_rasters[[as.character(input$year)]],
        colors = pal_rain,
        opacity = 0.85
      ) %>%
      addLegend(
        pal = pal_rain,
        values = all_vals,
        title = "Rainfall (mm)"
      )
  })

  observe({
    leafletProxy("rain_map") %>%
      clearImages() %>%
      addRasterImage(
        annual_rain_rasters[[as.character(input$year)]],
        colors = pal_rain,
        opacity = 0.85
      )
  })
}

shinyApp(ui, server)
```

Annual Rainfall Over Time

```{r}
# List of rainfall tifs
rain_base_path <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data"

all_tifs <- list.files(rain_base_path, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter years 1990-2025
rain_files <- all_tifs[grepl("_(199\\d|200\\d|201\\d|202[0-5])_\\d{2}\\.tif$", all_tifs)]
if(length(rain_files)==0) stop("No files found!")

# 2) Load rasters
rain_stack <- rast(rain_files)
names(rain_stack) <- gsub(".*_(\\d{4}_\\d{2})\\.tif$", "\\1", basename(rain_files))

# Extract year from layer names
years <- as.integer(substr(names(rain_stack), 1, 4))
unique_years <- sort(unique(years))

# Compute annual mean rainfall
annual_means <- map_dbl(unique_years, function(y){
  layers <- rain_stack[[which(years == y)]]
  vals <- values(layers)
  vals <- vals[!is.na(vals)]
  mean(vals)
})

mean_rain_per_year <- data.frame(
  year = unique_years,
  mean_rain_mm = annual_means
)

# Line Chart Plot
ggplot(mean_rain_per_year, aes(x=year, y=mean_rain_mm)) +
  geom_line(color="steelblue", size=1) +
  geom_point(color="darkred") +
  labs(
    title="Oʻahu Island-Wide Mean Rainfall (1990-2025)",
    x="Year",
    y="Mean Rainfall (mm)"
  ) +
  theme_minimal(base_size = 14)
```

## NDVI Data From 2000-2025

Loading and cleaning NDVI data

```{r}
# List all NDVI GeoTIFFs (Change this path to where you have the data stored)
ndvi_base <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data/oahu_ndvi"

all_ndvi_tifs <- list.files(ndvi_base, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter years
ndvi_files <- all_ndvi_tifs[
  grepl("ndvi_modis_day_oa_data_map_(200\\d|201\\d|202[0-4])_\\d{2}_\\d{2}\\.tif$", all_ndvi_tifs)
]

cat("Total NDVI files detected:", length(ndvi_files), "\n")

# List files that are able to be used

valid_ndvi <- list()

for(f in ndvi_files){
  try({
    r <- rast(f)
    valid_ndvi[[length(valid_ndvi) + 1]] <- r
  }, silent = TRUE)
}

cat("Successfully loaded:", length(valid_ndvi), "NDVI rasters\n")
cat("Unreadable files:", length(ndvi_files) - length(valid_ndvi), "\n")

# List bad files for transparency
bad_files <- setdiff(ndvi_files, sapply(valid_ndvi, sources))
if(length(bad_files) > 0){
  cat("Unreadable NDVI files:\n")
  print(bad_files)
}

if(length(valid_ndvi) == 0){
  stop("No valid NDVI rasters could be loaded.")
}

# Stack useable rasters
ndvi_stack <- rast(valid_ndvi)

names(ndvi_stack) <- gsub(
  ".*_(\\d{4}_\\d{2}_\\d{2})\\.tif$", 
  "\\1",
  basename(sources(ndvi_stack))
)

# Extract date components
dates <- names(ndvi_stack)
years <- as.integer(substr(dates, 1, 4))
months <- substr(dates, 1, 7)
```

Monthly and annual computations

```{r}
# Monthly mean(2000–2025)
monthly_ids <- unique(months)

monthly_ndvi <- map_dfr(monthly_ids, function(m){
  idx <- which(months == m)
  r <- mean(ndvi_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]
  
  data.frame(
    month = m,
    mean_ndvi = mean(vals),
    min_ndvi  = min(vals),
    max_ndvi  = max(vals),
    sd_ndvi   = sd(vals),
    p10_ndvi  = quantile(vals, 0.10),
    p90_ndvi  = quantile(vals, 0.90)
  )
})
head(monthly_ndvi)

# Annual NDVI (2000-2025)
unique_years <- sort(unique(years))

annual_ndvi <- map_dfr(unique_years, function(y){
  idx <- which(years == y)
  r <- mean(ndvi_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]
  
  data.frame(
    year = y,
    mean_ndvi = mean(vals)
  )
})
annual_ndvi
```

Long-Term Mean NDVI Map

```{r}
annual_mean_raster <- mean(ndvi_stack)
annual_mean_ndvi <- annual_mean_raster

annual_df <- as.data.frame(annual_mean_raster, xy=TRUE)
colnames(annual_df) <- c("x", "y", "ndvi")

ggplot(annual_df, aes(x=x, y=y, fill=ndvi)) +
  geom_raster() +
  scale_fill_gradientn(
    colors = (brewer.pal(9,"YlGn")),
    name="NDVI",
    guide = guide_colorbar(
      barwidth = 1.5, barheight = 15,
      title.position="top",
      title.hjust=0.5
    )
  ) +
  coord_equal() +
  labs(
    title="Oʻahu Mean NDVI (2000–2024)",
    subtitle="MODIS NDVI daily data aggregated to long-term mean",
    x="Longitude", 
    y="Latitude"
  ) +
  theme_minimal(base_size=14) +
  theme(
    legend.position="right",
    plot.title=element_text(hjust=0.5, face="bold"),
    plot.subtitle=element_text(hjust=0.5)
  )
```

Annual Mean NDVI 2000-2024 w/ Slider

```{r}
# Extract Years from File Names
years <- as.integer(gsub(
  ".*_(\\d{4})_\\d{2}_\\d{2}\\.tif$", 
  "\\1",
  basename(sources(ndvi_stack))
))

unique_years <- sort(unique(years))

# Annual NDVI Rasters
annual_ndvi_rasters <- list()
for(y in unique_years){
  idx <- which(years == y)
  annual_ndvi_rasters[[as.character(y)]] <- mean(ndvi_stack[[idx]])
}

# Color Palette
all_vals <- unlist(lapply(annual_ndvi_rasters, values))
pal <- colorNumeric(
  palette = brewer.pal(9, "YlGn"),
  domain = all_vals,
  na.color = "transparent"
)

# Shiny UI
ui <- fluidPage(
  titlePanel("Oʻahu Annual Mean NDVI"),
  sidebarLayout(
    sidebarPanel(
      sliderInput(
        "year",
        "Select Year:",
        min = min(unique_years),
        max = max(unique_years),
        value = min(unique_years),
        step = 1,
        sep = ""
      )
    ),
    mainPanel(
      leafletOutput("ndvi_map", height = "600px")
    )
  )
)

# Shiny Server
server <- function(input, output, session) {

  output$ndvi_map <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%
      addRasterImage(
        annual_ndvi_rasters[[as.character(input$year)]],
        colors = pal,
        opacity = 0.8
      ) %>%
      addLegend(
        pal = pal,
        values = all_vals,
        title = "NDVI"
      )
  })

  observe({
    leafletProxy("ndvi_map") %>%
      clearImages() %>%
      addRasterImage(
        annual_ndvi_rasters[[as.character(input$year)]],
        colors = pal,
        opacity = 0.8
      )
  })
}

shinyApp(ui, server)
```

Annual NDVI Time-Series Plot

```{r}
ggplot(annual_ndvi, aes(x=year, y=mean_ndvi)) +
  geom_line(color="forestgreen", linewidth=1) +
  geom_point(color="darkred") +
  labs(
    title="Oʻahu Annual Mean NDVI (2000–2024)",
    x="Year",
    y="Mean NDVI"
  ) +
  theme_minimal(base_size=14)
```

## Temperature Data From 1990-2025

Load and Clean Data

```{r}
# Load Temp Data
temp_base <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data/oahu_temp"

all_temp_tifs <- list.files(temp_base, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter years 1990-2025
temp_files <- all_temp_tifs[grepl("temperature_mean_month_oa_data_map_(199\\d|200\\d|201\\d|202[0-5])_\\d{2}\\.tif$", all_temp_tifs)]

cat("Total temperature files detected:", length(temp_files), "\n")

if(length(temp_files) == 0) stop("No temperature files found. Check folder or filenames!")

# Load valid rasters
valid_temp <- list()
for(f in temp_files){
  try({
    r <- rast(f)
    valid_temp[[length(valid_temp) + 1]] <- r
  }, silent = TRUE)
}

cat("Successfully loaded:", length(valid_temp), "temperature rasters\n")

# Stack all rasters
temp_stack <- rast(valid_temp)
names(temp_stack) <- gsub(".*_(\\d{4}_\\d{2})\\.tif$", "\\1", basename(temp_files))

# Monthly Metrics
months <- substr(names(temp_stack),1,7)

monthly_temp <- map_dfr(unique(months), function(m){
  idx <- which(months == m)
  r <- mean(temp_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]
  
  data.frame(
    month = m,
    mean_temp = mean(vals),
    min_temp  = min(vals),
    max_temp  = max(vals),
    sd_temp   = sd(vals),
    p10_temp  = quantile(vals,0.10),
    p90_temp  = quantile(vals,0.90)
  )
})

head(monthly_temp)

# Annual Mean Temperature Rasters
years <- as.integer(substr(names(temp_stack),1,4))
unique_years <- sort(unique(years))

annual_temp <- map_dfr(unique_years, function(y){
  idx <- which(years == y)
  r <- mean(temp_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]
  
  data.frame(
    year = y,
    mean_temp = mean(vals)
  )
})

annual_temp
```

Long-Term Mean Temperature

```{r}
annual_mean_raster <- mean(temp_stack)
annual_mean_temp <- mean(temp_stack)

annual_df <- as.data.frame(annual_mean_temp, xy=TRUE)
colnames(annual_df) <- c("x","y","temp")

ggplot(annual_df, aes(x=x, y=y, fill=temp)) +
  geom_raster() +
  scale_fill_gradientn(
    colors = rev(brewer.pal(9,"RdYlBu")),
    name="Temperature (°C)"
  ) +
  coord_equal() +
  labs(
    title="Oʻahu Long-Term Mean Temperature (1990–2025)",
    x="Longitude",
    y="Latitude"
  ) +
  theme_minimal(base_size=14) +
  theme(
    legend.position="right",
    plot.title=element_text(hjust=0.5, face="bold")
  )
```

Annual Mean Temperature 1990-2025 w/ Slider

```{r}
# Annual Mean Rasters
years <- as.integer(substr(names(temp_stack),1,4))
unique_years <- sort(unique(years))

annual_rasters <- list()
for(y in unique_years){
  idx <- which(years == y)
  r <- mean(temp_stack[[idx]])
  annual_rasters[[as.character(y)]] <- r
}

# Palette
all_vals <- unlist(lapply(annual_rasters, values))
pal <- colorNumeric(
  palette = rev(brewer.pal(9, "RdYlBu")),
  domain = all_vals,
  na.color = "transparent"
)

# Shiny App
ui <- fluidPage(
  titlePanel("Oʻahu Annual Mean Temperature"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("year",
                  "Select Year:",
                  min = min(unique_years),
                  max = max(unique_years),
                  value = min(unique_years),
                  step = 1,
                  sep = "")
    ),
    mainPanel(
      leafletOutput("temp_map", height = "600px")
    )
  )
)

server <- function(input, output, session) {
  
  output$temp_map <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%
      addRasterImage(
        annual_rasters[[as.character(input$year)]],
        colors = pal,
        opacity = 0.8
      ) %>%
      addLegend(
        pal = pal,
        values = all_vals,
        title = "Temperature (°C)"
      )
  })
  
# Update map when slider changes
  observe({
    leafletProxy("temp_map") %>%
      clearImages() %>%
      addRasterImage(
        annual_rasters[[as.character(input$year)]],
        colors = pal,
        opacity = 0.8
      )
  })
}

shinyApp(ui, server)
```

Annual Temperature Time-Series Plot

```{r}
ggplot(annual_temp, aes(x=year, y=mean_temp)) +
  geom_line(color="firebrick", linewidth=1) +
  geom_point(color="darkblue") +
  labs(
    title="Oʻahu Annual Mean Temperature (1990–2025)",
    x="Year",
    y="Mean Temperature (°C)"
  ) +
  theme_minimal(base_size=14)
```

## Watershed and Haole Koa

Load and clean data

```{r}
# Load watersheds and clip to just Oʻahu
ws_url <- "https://geodata.hawaii.gov/arcgis/rest/services/FreshWater/MapServer/23/query?where=1=1&outFields=*&f=geojson"
watersheds_all <- sf::read_sf(ws_url) %>% st_transform(4326)

# Oʻahu bounding box
oahu_box <- st_bbox(c(
  xmin = -158.3,
  xmax = -157.6,
  ymin = 21.2,
  ymax = 21.8
), crs = st_crs(4326)) %>% st_as_sfc()

# Clip to Oʻahu only
watersheds <- st_intersection(watersheds_all, oahu_box)

# Clean watershed name field
watersheds <- watersheds %>%
  rename(WATERSHED = name)

# Remove unnamed tiny slivers
watersheds <- watersheds %>% 
  filter(!grepl("Unnamed", WATERSHED, ignore.case = TRUE))

# Download Haole Koa from GBIF and clip to Oʻahu
hk_data <- occ_search(
  scientificName = "Leucaena leucocephala",
  country = "US",
  stateProvince = "Hawaii",
  hasCoordinate = TRUE,
  limit = 20000
)$data %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

hk_sf <- st_as_sf(
  hk_data,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326
)

# Clip ONLY Oʻahu HK points
hk_oahu <- st_intersection(hk_sf, oahu_box)

# Assign HK to watersheds and compute density
hk_joined <- st_join(
  hk_oahu,
  watersheds %>% select(WATERSHED),
  left = FALSE
)

hk_counts <- hk_joined %>%
  st_drop_geometry() %>%
  count(WATERSHED, name = "hk_points")

watersheds2 <- watersheds %>%
  left_join(hk_counts, by = "WATERSHED") %>%
  mutate(
    hk_points      = ifelse(is.na(hk_points), 0L, hk_points),
    area_km2       = as.numeric(st_area(geometry)) / 1e6,
    hk_density_km2 = hk_points / area_km2
  )
```

Haole Koa Occurences

```{r}
ggplot() +
  geom_sf(data = watersheds2, fill = NA, color = "gray40") +
  geom_sf(data = hk_oahu, color = "darkgreen", alpha = 0.6, size = 1) +
  coord_sf(xlim = c(-158.3, -157.6), ylim = c(21.2, 21.8)) +
  labs(
    title = "Haole Koa Occurrences (GBIF) on Oʻahu",
    subtitle = "Points clipped to Oʻahu watershed boundaries"
  ) +
  theme_minimal()
```

Haole Koa Density by Watershed

```{r}
ggplot(watersheds2) +
  geom_sf(aes(fill = hk_density_km2)) +
  scale_fill_viridis_c(
    option = "C",
    name = "HK Density (points/km²)",
    na.value = "gray90"
  ) +
  coord_sf(xlim = c(-158.3, -157.6), ylim = c(21.2, 21.8)) +
  labs(
    title = "Haole Koa Density by Watershed — Oʻahu"
  ) +
  theme_minimal()
```

Ranking Bar Chart

```{r}
topN <- 20

hk_summary <- watersheds2 %>%
  st_drop_geometry() %>%
  arrange(desc(hk_density_km2)) %>%
  slice_head(n = topN)

ggplot(hk_summary,
       aes(x = reorder(WATERSHED, hk_density_km2),
           y = hk_density_km2)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(
    title = paste("Top", topN, "Watersheds by Haole Koa Density"),
    x = "Watershed",
    y = "Density (points/km²)"
  ) +
  theme_minimal()

```

## Ignition Probability

Load Data

```{r}
ign_base <- "/Users/roronorazoro/Documents/GitHub/Oahu-Rainfall-Watershed-Risk/data/oahu_ignit"

all_ign_tifs <- list.files(ign_base, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)

# Filter valid filenames
ign_files <- all_ign_tifs[
  grepl("ignition_probability_day_lead00_oa_data_map_(200\\d|201\\d|202[0-5])_\\d{2}_\\d{2}\\.tif$", 
        all_ign_tifs)
]

length(ign_files)

# Load rasters
valid_ign <- list()
bad_ign <- character()

for (f in ign_files) {
  try({
    r <- rast(f)
    valid_ign[[length(valid_ign) + 1]] <- r
  }, silent = TRUE)
}

ign_stack <- rast(valid_ign)

# Extract dates
ign_dates <- gsub(".*_(\\d{4}_\\d{2}_\\d{2})\\.tif$", "\\1", basename(sources(ign_stack)))
ign_years <- as.integer(substr(ign_dates, 1, 4))
ign_months <- substr(ign_dates, 1, 7)

```

Calculations

```{r}
# Compute Monthly & Annual Ignition Probability

# Monthly mean (2002–2025)
monthly_ids <- unique(ign_months)

monthly_ign <- map_dfr(monthly_ids, function(m){
  idx <- which(ign_months == m)
  r <- mean(ign_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]

  tibble(
    month = m,
    mean_ign = mean(vals),
    p10_ign  = quantile(vals, 0.10),
    p90_ign  = quantile(vals, 0.90)
  )
})

# Annual ignition
unique_years_ign <- sort(unique(ign_years))

annual_ign <- map_dfr(unique_years_ign, function(y){
  idx <- which(ign_years == y)
  r <- mean(ign_stack[[idx]])
  vals <- values(r)
  vals <- vals[!is.na(vals)]

  tibble(
    year = y,
    mean_ign = mean(vals)
  )
})
```

Long-Term Ignition Probability

```{r}
# Long-term mean ignition probability
ign_mean_raster <- mean(ign_stack)

ign_df <- as.data.frame(ign_mean_raster, xy = TRUE)
colnames(ign_df) <- c("x", "y", "ignition")

ggplot(ign_df, aes(x=x, y=y, fill=ignition)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Ignition Prob.") +
  coord_equal() +
  labs(
    title="Oʻahu Long-Term Ignition Probability (2002–2025)"
  ) +
  theme_minimal()
```

Annual Mean Ignition Probability

```{r}
ggplot(annual_ign, aes(x=year, y=mean_ign)) +
  geom_line(color="firebrick", linewidth=1.2) +
  geom_point(color="black") +
  labs(
    title="Oʻahu Annual Mean Ignition Probability (2002–2025)",
    x="Year",
    y="Mean Ignition Probability"
  ) +
  theme_minimal()

```

Ignition Probablity by Watershed

```{r}
# Extract watershed polygons (from previous HK work)
ws <- watersheds

# Extract ignition mean for each watershed
ign_ws_values <- exactextractr::exact_extract(ign_mean_raster, ws, 'mean')

ws_ign <- ws %>%
  mutate(mean_ignition = ign_ws_values)

ggplot(ws_ign) +
  geom_sf(aes(fill = mean_ignition)) +
  scale_fill_viridis_c(name = "Mean Ignition\nProbability") +
  labs(
    title = "Ignition Probability by Watershed — Oʻahu"
  ) +
  theme_minimal()
```

Ranking Watershed by Ignition Probability (Top 20)

```{r}
ws_rank <- ws_ign %>%
  st_drop_geometry() %>%
  arrange(desc(mean_ignition)) %>%
  slice_head(n = 20)   # top 20 only

ggplot(ws_rank,
       aes(x=reorder(WATERSHED, mean_ignition),
           y=mean_ignition)) +
  geom_col(fill="firebrick") +
  coord_flip() +
  labs(
    title = "Top 20 Watersheds by Ignition Probability",
    x="Watershed",
    y="Ignition Probability"
  ) +
  theme_minimal()

```

## Machine Learning

Load and Clean Data

```{r}
#  MACHINE LEARNING: WATERSHED STRESS INDEX (WSI)

# Load Elevation Using Elevatr
oahu_bbox <- data.frame(
  x = c(-158.3, -157.6),
  y = c(21.2, 21.8)
)

elev_raster <- get_elev_raster(
  locations = oahu_bbox,
  prj = "+proj=longlat +datum=WGS84",
  z = 10,   
  expand = 0.05
)

elev_raster <- rast(elev_raster)

# Objects should already exist:
#   annual_mean        = rainfall long-term mean
#   annual_mean_ndvi   = NDVI long-term mean
#   annual_mean_temp   = temperature long-term mean
#   ign_mean_raster    = ignition long-term mean
#   watersheds         = Oʻahu watershed polygons
#   watersheds2        = watersheds + hk_density_km2

# Rainfall
ws_rain_values <- exact_extract(annual_mean, watersheds, 'mean')

# NDVI
ws_ndvi_values <- exact_extract(annual_mean_ndvi, watersheds, 'mean')

# Temperature
ws_temp_values <- exact_extract(annual_mean_temp, watersheds, 'mean')

# Ignition probability
ws_ign_values <- exact_extract(ign_mean_raster, watersheds, 'mean')

# Elevation
ws_elev_values <- exact_extract(elev_raster, watersheds, 'mean')
```

Machine Learning Dataframe and Training

```{r}
watershed_ml <- tibble(
  WATERSHED    = watersheds$WATERSHED,
  hk_density   = watersheds2$hk_density_km2,
  mean_rain    = ws_rain_values,
  mean_ndvi    = ws_ndvi_values,
  mean_temp    = ws_temp_values,
  mean_elev    = ws_elev_values,
  ignition_obs = ws_ign_values
) %>% drop_na()

# Random Forest to predict ignition

set.seed(123)

rf_model <- randomForest(
  ignition_obs ~ hk_density + mean_rain + mean_ndvi + mean_temp + mean_elev,
  data = watershed_ml,
  importance = TRUE,
  na.action = na.omit
)

print(rf_model)
importance(rf_model)

# Predict ignition and build WSI
watershed_ml$ign_pred <- predict(rf_model, watershed_ml)

# Scale helper
scale01 <- function(x) (x - min(x)) / (max(x) - min(x))

watershed_ml <- watershed_ml %>%
  mutate(
    rain_s   = scale01(mean_rain),
    ndvi_s   = scale01(mean_ndvi),
    temp_s   = scale01(mean_temp),
    elev_s   = scale01(mean_elev),
    hk_s     = scale01(hk_density),
    ign_s    = scale01(ign_pred)
  )

# Weighted WSI formula (0–1 scale, higher = more environmental stress)
watershed_ml$WSI <- (
  watershed_ml$hk_s * 0.30 +          # invasive pressure
  watershed_ml$temp_s * 0.20 +        # heat stress
  watershed_ml$ign_s  * 0.30 +        # predicted fire likelihood
  watershed_ml$elev_s * 0.10 +        # elevation dryness
  (1 - watershed_ml$ndvi_s) * 0.10    # low NDVI = vegetation stress
)
```

Map WSI Over Oʻahu

```{r}
ws_map <- watersheds %>%
  left_join(
    watershed_ml %>% select(WATERSHED, WSI),
    by = "WATERSHED"
  )

ggplot(ws_map) +
  geom_sf(aes(fill = WSI)) +
  scale_fill_viridis_c(
    option = "C",
    name = "Watershed\nStress Index",
    na.value = "gray90"
  ) +
  coord_sf(xlim = c(-158.3, -157.6), ylim = c(21.2, 21.8)) +
  labs(
    title = "Watershed Stress Index (WSI) — Oʻahu",
    subtitle = "Random Forest model integrating environmental factors and ignition probability"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, face = "bold"),   # title large
    plot.subtitle = element_text(size = 8)                 # subtitle half size
  )
```

Evaluation Metrics

```{r}
# Out-of-bag error (RF internal validation)
oob_error <- rf_model$mse[length(rf_model$mse)]  # last MSE value
oob_rmse <- sqrt(oob_error)

cat("---- Random Forest Model Performance ----\n")
cat("OOB MSE:  ", round(oob_error, 4), "\n")
cat("OOB RMSE: ", round(oob_rmse, 4), "\n\n")

# R-squared (based on actual vs predicted)
actual <- watershed_ml$ignition_obs
pred   <- watershed_ml$ign_pred

# R-squared
rss <- sum((actual - pred)^2)
tss <- sum((actual - mean(actual))^2)
r2  <- 1 - rss/tss

# MAE
mae <- mean(abs(actual - pred))

# RMSE
rmse <- sqrt(mean((actual - pred)^2))

cat("R-squared: ", round(r2, 4), "\n")
cat("MAE:       ", round(mae, 4), "\n")
cat("RMSE:      ", round(rmse, 4), "\n")
```

Feature Importance Plot

```{r}
importance_df <- data.frame(
  Variable = rownames(importance(rf_model)),
  Importance = importance(rf_model)[, 1]
)

ggplot(importance_df, aes(x=reorder(Variable, Importance), y=Importance)) +
  geom_col(fill="steelblue") +
  coord_flip() +
  labs(
    title = "Random Forest Variable Importance",
    x = "Predictor",
    y = "Importance (IncMSE)"
  ) +
  theme_minimal()
```
